Experience Report: Haskell in Computational Biology

### 2012-03-11 mrfy_experience_report.tex
new story about memoization

Given the \texttt{Scored} abstraction, we can easily implement
Viterbi's equations for a match state.
To~implement the equation at the top right of \figref{viterbi},
we~choose the best solution
from among the three predecessor states;
we~use~\texttt{/+/} to add transition and emission scores,
and we use \texttt{fmap} to add \texttt{Mat} to the list of states
encountered up to this point:
\remark{TODO: 
(1)~Call the inner function \texttt{vee}
(2)~Use \texttt{minimum (map from [Mat, Ins, Del])}
(3)~Explain the recursive call
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\smallverbatiminput{vfix}
Functions \texttt{tProb} and \texttt{edge} combine to select the
appropriate transition score (one of the $a'$~values).
Implementing the equation in this form was not only intellectually
satisfying; it~also helped us find
bugs.\remark{NR doesn't think the bug story is that interesting.}
The most memorable bug was that in one of our base cases, 
we~mistakenly returned an empty path of states;
the correct result was a singleton path containing the initial state.
Very quickly after observing a missing state in the output, we were
able to find the faulty case in the code.

Whatever the attractions of the code above, our work also
requires high performance.
To make the simple recursive code efficient, we used Luke Palmer's
\texttt{Data.Memocombinators} package.
The function \verb+viterbi'+ is unmemoized,
and it makes recursive calls to \verb+viterbi+, which is memoized:
\smallverbatiminput{memo}
The value
\texttt{numNodes} represents~$\alignwidth$,
and \texttt{seqlen} represents~$N$.\remark{I hope this is right. ---NR}

This memoized recursive code performs just as well as the classic
dynamic-programming code.
And~the call to \texttt{Memo.memo3} is the \emph{only} part of the code
devoted to improving performance.
By~contrast, standard implementations of the Viterbi algorithm, such as in HMMER,
spend much of their code 
managing dynamic-programming tables.
Haskell enabled us write simple, performant code with little effort.
%
Because the memoized version so faithfully resembles the equations in
\figref{viterbi}, we~retired the classic dynamic-programming version.

@@

We hope you find the search code simple and straightforward:
\smallfuzzverbatiminput{11pt}{search}

@@

>>>>

Given the \texttt{Scored} abstraction, we can easily implement
Viterbi's equations for a match state.
To~implement the equation at the top right of \figref{viterbi},
we~choose the best solution
from among the three predecessor states;
we~use~\texttt{/+/} to add transition and emission scores,
and we use \texttt{fmap} to add \texttt{Mat} to the list of states
encountered up to this point:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\smallverbatiminput{vfix}
Functions \texttt{tProb} and \texttt{edge} combine to select the
appropriate transition score (one of the $a'$~values).
Function~\verb+vee''+ is the \emph{memoized} version of~\verb+vee'+,
so
the call to~\verb+vee''+ produces the same result as a recursive call
to~\verb+vee'+, but faster:
\smallverbatiminput{memo}
Functions \texttt{Memo.memo3} and \texttt{Memo.arrayRange} come from
Luke Palmer's
\path{Data.Memocombinators} package.
The value
\texttt{numNodes} represents~$\alignwidth$,
and \texttt{seqlen} represents~$N$.

Memoization makes \verb+vee'+ perform just as well as the classic
dynamic-programming code.
And~the call to \texttt{Memo.memo3} is the \emph{only} part of the code
devoted to improving performance.
By~contrast, standard implementations of the Viterbi algorithm, such as in HMMER,
spend much of their code 
managing dynamic-programming tables.
Haskell enabled us write simple, performant code with little effort.
%
Because the memoized version so faithfully resembles the equations in
\figref{viterbi}, we~retired the classic dynamic-programming version.

@@

We hope you find the search simple and straightforward:
\smallfuzzverbatiminput{11pt}{search}

@@

Implementing the equation in this form was not only intellectually
satisfying; it~also helped us find
bugs.
The most memorable bug was that in one of our base cases, 
we~mistakenly returned an empty path of states;
the correct result was a singleton path containing the initial state.
Very quickly after observing a missing state in the output, we were
able to find the faulty case in the code.
